The negative external effects of deforestation on downstream populations in Brazil

### 2024-08-14 output/documents/thesis_paper/paper.qmd
Finished main part of writing

---
title: "The negative external effects of deforestation on downstream populations in Brazil"
subtitle: "Master Thesis"
author: "Felix Schulz"
format:
  pdf:
    papersize: a4
    keep-tex: true
    toc: false
    number-sections: true
    link-citations: true
    include-in-header:  
      - text: |
          \usepackage{graphicx}
          \usepackage{amsfonts}
          \usepackage{amsmath}
          \usepackage{pifont}
          \usepackage{booktabs}
          \usepackage{dcolumn}
          \usepackage{makecell}
          \usepackage{threeparttable}
bibliography: "./Master Thesis.bib"
---

@@

In addition to the global reaching effects of deforestation on the climate, there is a substantial body of literature discussing its impact on local utility in the form of health outcomes. Empirical evidence from medical and geoscientific research find negative effects of deforestation-related wildfires and on malaria prevalence [see for example @reddingtonAirQualityHuman2015; @tuckerlimaDoesDeforestationPromote2017]. Given the existing evidence it appears likely that these dangers to the local communities are not considered to the extent it would be socially desireable [@NBERw31410].

@@

The impact of deforestation on health outcomes has been subject of several publications [@ellwangerDiversityLossClimate2020; @reddingtonAirQualityHuman2015]. The costs of soil erosion were also discussed in a broad range of literature [see for example @tellesCostsSoilErosion2011]. @NBERw31410 provide an excellent review on the economics of deforestation. Soil erosion as a driver of pollution and pollution-induced health costs has to the best of knowledge only been covered in a small-scale fashion by @maHumanHealthRisk2014 and therefore remains a gap in the research.

@@

Driven by short-run benefits of lumber or alternative land use, researchers observe unprecedented levels of rainforest degradation in the anthropocene [@malhiClimateChangeDeforestation2008]. This is of elevated importance as any loss of primary forests cannot be undone in any reasonable temporal scope [@shuklaAmazonDeforestationClimate1990; @NBERw31410]. The scientific evidence for the negative impacts of deforestation raises the concern that these externalities are not internalized by the forest extraction market and that deforestation does not occur at a socially optimal rate. At the core of this thought lies the (monetary) assessment of the societal cost of deforestation against the economical gains from alternative use and timber.

Many economic models have been developed to weigh these benefits and external costs on a local and global scale. In their analysis, @NBERw31410 describe several such models of decision making for deforestation. Their benchmark model, which they claim to closest resemble the microeconomic process of decisionmaking today, disregards externalities but instead focuses on short-run benefits. In fact, discounting of future income leads to a situation where this benchmark model does not guarantee extraction rates that are at or below a *maximum sustained yield* of forest regrowth. They note that lack of property rights and regulation further increases discounting rates in regions like Brazil.

Extended models then acknowledge the fact that deforestation is associated with social costs that are not internal to the parties with property rights to the forest [@balboniAppendixEconomicsTropical2023]. By modeling this cost as a fixed shift to the private marginal cost, the socially optimal equilibrium lies below the market-achieved private rate. This underlines the necessity to internalize these costs through policy. Economic theory yields several instruments to achieve this goal. Pigouvian taxes on logging or subsidies for landowner to *not* remove a forest are among the popular [@NBERw31410]. I discuss several such attempts in @sec-policy. It does however remain essential to know the extent of the social costs of deforestation for any such policy to work. My proposed paper therefore has important policy implications by shedding a light on the superlocal health cost of logging across Brazil.

@@

Brazil has since made extensive pledges to reduce deforestation-related emissions and implements a range of policy and monitoring instruments to ensure attainment [@tacconiLawEnforcementDeforestation2019]. Since its inception in 2008, the Amazon Fund is important cornerstone policy in Brazil's strategy against deforestation [@correaAmazonFund102019]. It is at its core an implementation of the principles of the Reducing Emissions from Deforestation and Forest Degradation (REDD+) framework. International donations are used to fund conservation projects in monitoring existing regulation, maintaining protected areas, supporting sustainable production and investments in innovations. The central idea is to make payments conditional on environmental efforts and thereby subsidize abatement for those who would not have been able to afford it void the transfers [@NBERw31410]. The effectiveness and efficiency of this command-and-control approach is however disputed [@correaAmazonFund102019]. *Bolsa Verde*, another example of a REDD+ program, paying poor landowners in the Amazon to maintain forest cover is considered to be cost effective [@RePEc:cgs:wpaper:110].

@@

Exogeneity requires the instrument to neither directly (*exogeneity*) nor indirectly influence health outcomes through any other variable than deforestation (*exclusion restriction*). The most obvious concern is that increased cloud coverage comes in hand with weather conditions that have effects on public health. For example, increased rainfall may lead to a rise in waterborne diseases or vector-borne diseases like malaria and dengue fever, which are sensitive to changes in precipitation. The cumulative nature of my outcome variable helps alleviate this concern. Since the independent variable of interest is upstream deforestation, I aggregate deforestation in an area of up to 1000 kilometers in river length. The larger the distance, the less likely weather is going to impact health outcomes at a given municipality. Nevertheless, I do include weather controls, namely average precipitation and temperature in my models.

In @sec-cloud_cover-deforestation I estimate a difference-in-difference-in-differences (DDD) regression of deforestation on cloud cover that also serves as a placebo test to the main specification in this section. Finally, another concern is that cloud coverage may be reversely causated by deforestation. Over the past few decades, numerical simulations predict a decrease in cloud cover in some parts of the Amazon, particularly in regions experiencing extensive deforestation [@royMesoscaleVegetationAtmosphere2009]. This analysis is based on the theoretical finding that deforestation disrupts local hydrological cycles, reducing evapotranspiration and subsequent cloud formation [@lejeuneInfluenceAmazonianDeforestation2015]. On the contrary, empirical evidence from the Brazilian rainforest contradicts this, showing that deforestation is associated with increases in forest cover [@royMesoscaleVegetationAtmosphere2009, @xuContrastingImpactsForests2022]. In any case, there appears to be a theoretical and empirical link between cloud cover and deforestation. I am not able to rule out this reverse causality between the instrument and the endogenous regressor. In comparison to @assuncaoDETERingDeforestationAmazon2023, this is a more elevated issue in my work because I use cloud cover to directly instrument deforestation opposed to law enforcement activity. Any further findings should therefore be read keeping this in mind.

My main analysis then follows the following model structure. I attempt to estimate the causal relationship between deforestation and health outcomes with a two-stage least squares (TSLS) approach. In the first stage I regress the potentially endogenous regressor deforestation on the instrument, cloud coverage, and any control variables. Any variation in deforestation remaining after recovering the fitted values from this model should then be exogenous i.r.t. the main relationship in my model. The first stage formula is given by the equation:

@@

@@

For my research I source and combine a selection of public data from the realms of environmental science, health and economics. For replication of data generation, please see section @sec-appendix-data. Not all data is available in all time periods. In figureFor my central analyis of deforestation and health outcomes, I subset data for which a complete panel without missing values is available. The municipalities included in these panels are shown in @ .

@@

In the field of land use classification there have been made several recent advancements allowing for detailed and thorough analyses of changes of land cover and their implications. For Brazil specifically, the MapBiomas project [@souzaReconstructingThreeDecades2020] provides detailed, annual land cover classification at 30m resolution. I obtain *Collection 8* rasters and aggregate the data by drainage polygon and year. To this end, I simplify the data's classification and re-group them into the categories of forest, pasture, agriculture, urban, and mining.^[See X for a correspondence table between MapBiomas and my classes.] Finally, I count deforestation events as pixel-wise changes between two subsequent years from forest to any other class.

@@

In my analysis I observe health outcomes from two major data sources: I obtain death records from @ministeriodasaudedobrasilmsSistemaInformacoesMortalidade2022. Upon the passing of every Brazilian, medical professionals are required to fill out a standardized form that is collected by the responsible government agency. Each record states the age and home municipality of the diseased. Aggregating by municipality and combining it with birth and population estimates from @ministeriodasaudedobrasilmsNascidosVivosSINASC2023 and @institutobrasileirodegeografiaeestatisticaibgePopulacaoBrasileira2022 respectively allows me to calculate mortality rates.

@@

The DETER project [@dinizDETERBNewAmazon2015] publishes monthly cloud polygons in which they were not able to observe land cover changes. With some sporadic gaps, this data is mostly complete from mid-2005 until 2017. I rasterize and aggregate the data by municipality and year. To allow for analyses before 2005 and outside the project's boundaries (the legal Amazon), I also access CLARA-A3, a sensor-based long-term collection of cloud cover data by @copernicusclimatechangeserviceCloudPropertiesGlobal2022. It is important to note that the definition of cloud cover in both sources differs significantly. While CLARA report average monthly cloud cover in a given cell throughout the month, DETER reports areas in which they were not able to make any observations within the same time frame. Also, while both sensors provide great temporal detail of two or one daily observation respectively, the exact times of observation and spatial resolution (1.1km and 250m) differ between the two [@noaanationalcentersforenvironmentalinformationAdvancedVeryHigh1978, @nasaMODISSpecifications2024]. In @fig-stations I provide example images of cloud cover for both sources and show that the resulting municipal indicators are correlated.

@@

In order to trace potential outcomes down the flow of rivers, I compute a detailed model of all rivers in Brazil. Based on @sistemanacionaldeinformacoessobrerecursoshidricossnirhBHO20175KTrecho2017 and using tools of network analysis, I identify 1025 sub-networks identified by their shared estuary into the ocean, 123,140 confluences between rivers, and 125,192 source points. For every river segment I calculate its distance for the estuary to the sea. For further processing I split rivers at municipality borders to make both bijectively relatable. This network model is then used to calculate upstream reachability information for every node, which I aggregate by municipality. In @ I provide an example of the detail of this data.

My research design also requires me to link deforestation to rivers. Any sediments rivers would follow the flow of water through the terrain. Starting from drainage trenches, the flow of water accumulates until the stream can be considered a river. Surrounding any river there is therefore a watershed area which feeds into it. To model this, I split rivers into segments of 1km and then use the methodology of @bartos_2020 with the digital elevation model provided by @europeanspaceagencyCopernicus2022 to calculate exact drainage polygons around the river segments. This approach requires extensive data pre-processing and matching algorithms which I further elaborate on in @sec-appendix-hydro. These polygons are then used to extract deforestation as discussed above. In @ I present an example of the detailed results of this process.

@@

The Brazilian National Water Agency (ANA) collects standardized information on water quality in the @agencianacionaldeaguasanaPortalQualidadeAguas2024. I download all available historical data from the HidroWeb system of the National Water Resources Information System (SNIRH) and aggregate them by year, resulting in an unbalanced panel of 4,522 stations in 45,094 data points. In my analysis I focus on the variables of pH, turbidity, biochemical oxygen demand (BOD5.20), dissolved oxygen, total residue, total nitrogen, and nitrate concentration. Not all variables are measured at all times, please refer to @tbl-summary-sensor-data for summary statistics. A map of the spatial distribution of all stations is provided in @fig-stations.

@@

To account for potential confounding factors, several control variables are included in the analysis. These controls are GDP per capita, average schooling grades, a vaccination index, primary care coverage, doctors per capita, urbanization and the availability of sewage systems. Data for GDP per capita are sourced from @institutobrasileirodegeografiaeestatisticaibgeProdutoInternoBruto1996. Average schooling grades in form of the IDEB index are sourced @institutonacionaldeestudosepesquisaseducacionaisinepIndiceDesenvolvimentoEducacao2005. The vaccination index, representing the coverage of essential immunizations, is derived from health department records @ministeriodasaudemsSistemaImunizacoes1994.^[I focus on the number of administered vaccinations against yellow fever, hepatitis A and B, polio, and Haemophilus influenzae type B. In the analysis I use a rolling 5-year average of annual vaccination rates of newborns.] I obtain the primary care coverage and the number of doctors per capita from @ministeriodasaudemsCoberturaAtencaoBasica2007. Finally, data on urbanization and sewage availability are obtained from municipal health and infrastructure reports @ministeriododesenvolvimentoregionalmdrSistemaNacionalInformacoes1995.

@@

If deforestation was to expose populations to health outcomes along rivers, one should be able to estimate an impact on sensor-measured river pollution. Evidence from environmental science shows that soil erosion increases turbidity by introducing sediments into water bodies, reducing light penetration and affecting aquatic life [@unitedstatesenvironmentalprotectionagencyepaTurbidityParameterFactsheet2021]. Eroded soils carry organic matter and nutrients, raising biochemical oxygen demand levels and potentially depleting dissolved oxygen, crucial for aquatic organisms [@unitedstatesenvironmentalprotectionagencyepaDissolvedOxygenParameter2021]. The increased runoff also contributes to higher total residues, nitrogen, and nitrate concentrations, leading to eutrophication and harmful algal blooms [@unitedstatesenvironmentalprotectionagencyepaNutrientsParameterFactsheet2021]. Altered pH levels can further disrupt the ecological balance and water quality [@unitedstatesenvironmentalprotectionagencyepaPHParameterFactsheet2021].

In @tbl-reg-deforestation-sensors I present regression results for the corellation between aggregate upstream deforestation and pollution. I find that an increase of deforestation by one percentage point is associated with a decrease in the water's pH level by 0.02 or -0.03 standard deviations, an increase in turbidity by 0.62 or 0.03 standard deviations and a decrease of the nitrate level by 0.04 or -0.06 standard deviations c.p.. These coefficients are statistically significant at the 1%, 5% and 1% level respectively. The increase in turbidity matches expectations while the decrease in average nitrate levels is surprising.

@@

To further shed a light on the mechanism of impact behind this correlation I now turn to the effect of the different replacing land cover types. For any deforestation event, I identify what trees were replaced by and group this into four categories: Pasture, agriculture, urban environments, and mining. In @tbl-reg-deforestation-sensors-replacement I present OLS results of this analysis, with otherwise equivalent model parameters to the above. I find that deforestation for pasture and agriculture are both similarily correlated with pollution. On average, if upstream areas are replaced by these two land cover categories, pH and turbidity levels are increased. For agriculture I also find that the amount of dissolved oxygen is significantly lower. Urban expansion is associated with decreased levels of turbidity and total residue, as well as increased levels of nitrogen. Finally, mining is shown to have the most strong corellation with water pollution, with one p.p. increase in mining replacement being associated with a significant dcrease in pH and a strong increase in turbidity, total residue and nitrate readings.

@@

Surprisingly, nitrogen readings are on average lower when there is increased farming activity upstream. Also, deforestation for pasture is negatively correlated with biochemical oxygen demand and total residue levels are on average lower when forest is replaced by agriculture. These results do not match the expected outcomes from environmental theory and raise concerns to the validity of my research design. I therefore want to stress that I do not claim any of the results in this section to be causal relationships. In @tbl-reg-deforestation-sensors-replacement-levels I present results when regressing on levels of land cover instead of deforestation events. Finally, in @ I observe how the effect of deforestation differs with the distance to the deforestation.

## Cloud Coverage and Deforestation {#sec-cloud_cover-deforestation}

@@

Before I turn to the results of the analysis, I first present results in favor of the relevancy of the instrument. In @tbl-reg-deter-first-stage I regress upstream DETER cloud coverage on upstream deforestation for all municipalities. I present four different specifications, each consecutively adding control variables. Since data on doctors and primary care coverage is only available from 2010 onwards, the last model is estimated on a restricted sample. I find that in all specifications cloud coverage is positively correlated with deforestation. A 10 percentage point increase in cloud coverage is associated with an increase in deforestation of between 1.2 and 2.4 percentage points. I also report first stage F-statistics below and find that all statistics are above the @stockTestingWeakInstruments2005 threshold of 10.

{{< include /tables/reg_deforestation_DETER_first_stage.tex >}}

If my identification strategy holds, any regression estimates that utilize the instrumented deforestation rate should retrieve the true, causal relationship between deforestation and

{{< include /tables/reg_deforestation_DETER_mortality_tot.tex >}}

{{< include /tables/reg_deforestation_DETER_mortality_l1.tex >}}

>>>>

---
title: "The negative external effects of deforestation on downstream populations in Brazil"
subtitle: "Master Thesis"
author: "Felix Schulz"
format:
  pdf:
    papersize: a4
    keep-tex: true
    toc: false
    number-sections: true
    link-citations: true
    cite-method: biblatex
    biblatexoptions: style=apa
    include-in-header:  
      - text: |
          \usepackage{graphicx}
          \usepackage{amsfonts}
          \usepackage{amsmath}
          \usepackage{pifont}
          \usepackage{booktabs}
          \usepackage{dcolumn}
          \usepackage{makecell}
          \usepackage{threeparttable}
bibliography: "./Master Thesis.bib"
---

@@

In addition to the global reaching effects of deforestation on the climate, there is a substantial body of literature discussing its impact on local utility in the form of health outcomes. Empirical evidence from medical and geoscientific research find negative effects of deforestation-related wildfires and on malaria prevalence [see for example @reddingtonAirQualityHuman2015; @tuckerlimaDoesDeforestationPromote2017]. Given the existing evidence it appears likely that these dangers to the local communities are not considered to the extent it would be socially desireable [@balboniEconomicsTropicalDeforestation2023].

@@

The impact of deforestation on health outcomes has been subject of several publications [@ellwangerDiversityLossClimate2020; @reddingtonAirQualityHuman2015]. The costs of soil erosion were also discussed in a broad range of literature [see for example @tellesCostsSoilErosion2011]. @balboniEconomicsTropicalDeforestation2023 provide an excellent review on the economics of deforestation. Soil erosion as a driver of pollution and pollution-induced health costs has to the best of knowledge only been covered in a small-scale fashion by @maHumanHealthRisk2014 and therefore remains a gap in the research.

@@

Driven by short-run benefits of lumber or alternative land use, researchers observe unprecedented levels of rainforest degradation in the anthropocene [@malhiClimateChangeDeforestation2008]. This is of elevated importance as any loss of primary forests cannot be undone in any reasonable temporal scope [@shuklaAmazonDeforestationClimate1990; @balboniEconomicsTropicalDeforestation2023]. The scientific evidence for the negative impacts of deforestation raises the concern that these externalities are not internalized by the forest extraction market and that deforestation does not occur at a socially optimal rate. At the core of this thought lies the (monetary) assessment of the societal cost of deforestation against the economical gains from alternative use and timber.

Many economic models have been developed to weigh these benefits and external costs on a local and global scale. In their analysis, @balboniEconomicsTropicalDeforestation2023 describe several such models of decision making for deforestation. Their benchmark model, which they claim to closest resemble the microeconomic process of decisionmaking today, disregards externalities but instead focuses on short-run benefits. In fact, discounting of future income leads to a situation where this benchmark model does not guarantee extraction rates that are at or below a *maximum sustained yield* of forest regrowth. They note that lack of property rights and regulation further increases discounting rates in regions like Brazil.

Extended models then acknowledge the fact that deforestation is associated with social costs that are not internal to the parties with property rights to the forest [@balboniAppendixEconomicsTropical2023]. By modeling this cost as a fixed shift to the private marginal cost, the socially optimal equilibrium lies below the market-achieved private rate. This underlines the necessity to internalize these costs through policy. Economic theory yields several instruments to achieve this goal. Pigouvian taxes on logging or subsidies for landowner to *not* remove a forest are among the popular [@balboniEconomicsTropicalDeforestation2023]. I discuss several such attempts in @sec-policy. It does however remain essential to know the extent of the social costs of deforestation for any such policy to work. My proposed paper therefore has important policy implications by shedding a light on the superlocal health cost of logging across Brazil.

@@

Brazil has since made extensive pledges to reduce deforestation-related emissions and implements a range of policy and monitoring instruments to ensure attainment [@tacconiLawEnforcementDeforestation2019]. Since its inception in 2008, the Amazon Fund is important cornerstone policy in Brazil's strategy against deforestation [@correaAmazonFund102019]. It is at its core an implementation of the principles of the Reducing Emissions from Deforestation and Forest Degradation (REDD+) framework. International donations are used to fund conservation projects in monitoring existing regulation, maintaining protected areas, supporting sustainable production and investments in innovations. The central idea is to make payments conditional on environmental efforts and thereby subsidize abatement for those who would not have been able to afford it void the transfers [@balboniEconomicsTropicalDeforestation2023]. The effectiveness and efficiency of this command-and-control approach is however disputed [@correaAmazonFund102019]. *Bolsa Verde*, another example of a REDD+ program, paying poor landowners in the Amazon to maintain forest cover is considered to be cost effective [@RePEc:cgs:wpaper:110].

@@

Exogeneity requires the instrument to neither directly (*exogeneity*) nor indirectly influence health outcomes through any other variable than deforestation (*exclusion restriction*). The most obvious concern is that increased cloud coverage comes in hand with weather conditions that have effects on public health. For example, increased rainfall may lead to a rise in waterborne diseases or vector-borne diseases like malaria and dengue fever, which are sensitive to changes in precipitation. The spatially cumulative nature of my outcome variable helps alleviate this concern. Since the independent variable of interest is upstream deforestation, I aggregate deforestation in an area of up to 1000 kilometers in river length. The larger the distance, the less likely weather is going to impact health outcomes at a given municipality. Nevertheless, I do include weather controls, namely average precipitation and temperature in my models.

In addition to these theoretical arguments, I provide a placebo test for the validity of the instrument in @tbl-reg-placebo. Using an alternative data source on cloud cover, I test whether the latter predicts deforestation before DETER was initiated and in areas outside the legal Amazon where it is active. Since DETER only reports cloud cover for the areas and the duration in which it is active, I obtain data on cloud cover from ESA and aggregate them by municipality as discussed above. Importantly, both cloud cover indicators are constructed differently: While DETER reports the share of the municipality in which no observation was possible within a year, ESA utilized the higher spatial detail of their satellite data to calculate the overall share of cloud cover within the municipality and year. To show that both are approximately equivalent, I first regress the DETER on the ESA data and find that the former is explained well by the latter. This allows me to estimate the placebo test using the ESA in stead of the DETER data. In the second column I test whether cloud cover predicts deforestation rates in- and outside the legal Amazon before DETER was active. Encouragingly, none of the coefficients are statistically significant. After 2005, when DETER was initiated, cloud cover only predicts deforestation in the legal Amazon. This finding is statistically significant at all conventional levels. The instrument therefore passes my placebo test.

{{< include /tables/reg_placebo.tex >}}

Finally, a concern is that cloud coverage may be reversely causated by deforestation. Over the past few decades, numerical simulations predict a decrease in cloud cover in some parts of the Amazon, particularly in regions experiencing extensive deforestation [@royMesoscaleVegetationAtmosphere2009]. This analysis is based on the theoretical finding that deforestation disrupts local hydrological cycles, reducing evapotranspiration and subsequent cloud formation [@lejeuneInfluenceAmazonianDeforestation2015]. On the contrary, empirical evidence from the Brazilian rainforest contradicts this, showing that deforestation is associated with increases in forest cover [@royMesoscaleVegetationAtmosphere2009, @xuContrastingImpactsForests2022]. In any case, there appears to be a theoretical and empirical link between cloud cover and deforestation. I am not able to rule out this reverse causality between the instrument and the endogenous regressor. In comparison to @assuncaoDETERingDeforestationAmazon2023, this is a more elevated issue in my work because I use cloud cover to directly instrument deforestation opposed to law enforcement activity. Any further findings should therefore be read keeping this in mind.

My main analysis then follows the following model structure. I attempt to estimate the causal relationship between deforestation and health outcomes with a two-stage least squares (TSLS) approach. In the first stage I regress the potentially endogenous regressor of deforestation on the instrument, cloud coverage, and any control variables. Any variation in deforestation remaining after recovering the fitted values from this model should then be exogenous i.r.t. the main relationship in my model. The first stage formula is given by the equation:

@@

Given the spatial nature of my panel, there is a concern that spatial autocorrelation in the independent variables between municipalities may lead to biased or inefficient estimates in the regression models. Nearby municipalities cloud be subject to common shocks and may share characteristics. This violates the assumption of independence required for OLS analysis. If there is spatial dependency, standard errors are likely to be underestimated, leading to an overstatement of the significance of the predictors. As an extension to my main approach, following @salima2018spatial, I address this issue with a spatial error model as implemented in R by @millo2012splm. This entails obtaining within variation by demeaning and then iteratively obtained a maximum likelihood result for the model parameters. Feasible generalized least squares estimation yields estimates for $\beta$, model residuals, and the error variance $\sigma^2_{\epsilon}$. Using these best guesses, the model log-likelihood is maximized, yielding the spatial autoregressive parameter $\hat{\rho}$. The process is initiated with  $\hat{\rho}$ from OLS. This two-step (fGLS followed my ML) approach is repeated until convergence of the parameters. If spatial autocorrelation is present in the data, this method provides more accurate and reliable estimates.^[Please refer to @millo2012splm [p. 12] for details on likelihood functions.]

@@

For my research I source and combine a selection of public data from the realms of environmental science, health and economics. In @fig-data-workflow I present the entire workflow of the project to give an orientation of the different scripts and data sources used. Not all data is available in all time periods. In @fig-variable-temporal-ranges I present the availability of variables across time. For my central analyis of deforestation and health outcomes, I subset data for which a complete panel without missing values is available. The municipalities included in these panels are shown in @fig-balanced-panels.

@@

In the field of land use classification there have been made several recent advancements allowing for detailed and thorough analyses of changes of land cover. For Brazil specifically, the MapBiomas project [@souzaReconstructingThreeDecades2020] provides detailed, annual land cover classification at 30m resolution. I obtain *Collection 8* rasters and aggregate the data by drainage polygon and year. To this end, I simplify the data's classification and re-group them into the categories of forest, pasture, agriculture, urban, and mining.^[See @tbl-mapbiomas-equivalence for a correspondence table between MapBiomas and my classes.] Finally, I count deforestation events as pixel-wise changes between two subsequent years from forest to any other class.

In 1986, I classify about 68.5% of area in Brazil as forests. By 2020, this had reduced by 9.9 percentage points to 58.6%. In the legal Amazon, forest cover went from 84.5% to 72.6%. During that time and in the legal Amazon, area used for pasture strongly increases from 4.5% to 14.5%. Agricultural usage increases from 0.7% to 4% and cities make up about about 1.2% of area at the end of the timeframe. Mining is performed on 0.05% of area, an increase of 600% over the value in 1986.

@@

In my analysis I observe health outcomes from two major data sources: I obtain death records from @ministeriodasaudedobrasilmsSistemaInformacoesMortalidade2022. Upon the passing of every Brazilian, medical professionals are required to fill out a standardized form that is collected by the responsible government agency. Each record states the age and home municipality of the diseased. Aggregating by municipality and combining it with birth and population estimates from @ministeriodasaudedobrasilmsNascidosVivosSINASC2023 and @ibgePopulacaoBrasileira2022 respectively allows me to calculate mortality rates.

@@

The DETER project [@dinizDETERBNewAmazon2015] publishes monthly cloud polygons in which they were not able to observe land cover changes. With some sporadic gaps, this data is mostly complete from mid-2005 until 2017. I rasterize and aggregate the data by municipality and year. To allow for analyses before 2005 and outside the project's boundaries (the legal Amazon), I also access CLARA-A3, a sensor-based long-term collection of cloud cover data by @copernicusclimatechangeserviceCloudPropertiesGlobal2022. It is important to note that the definition of cloud cover in both sources differs significantly. While CLARA report average monthly cloud cover in a given cell throughout the month, DETER reports areas in which they were not able to make any observations within the same time frame. Also, while both sensors provide great temporal detail of two or one daily observation respectively, the exact times of observation and spatial resolution (1.1km and 250m) differ between the two [@noaanationalcentersforenvironmentalinformationAdvancedVeryHigh1978; @nasaMODISSpecifications2024]. In @fig-stations I provide example images of cloud cover for both sources and show that the resulting municipal indicators are correlated.

@@

In order to trace potential outcomes down the flow of rivers, I compute a detailed model of all rivers in Brazil. Based on @agencianacionaldeaguasesaneamentobasicoanaSistemaNacionalInformacoes2017 and using tools of network analysis, I identify 1025 sub-networks identified by their shared estuary into the ocean, 123,140 confluences between rivers, and 125,192 source points. For every river segment I calculate its distance for the estuary to the sea. For further processing I split rivers at municipality borders to make both bijectively relatable. This network model is then used to calculate upstream reachability information for every node, which I aggregate by municipality. In @fig-river-network-model I provide an example of the detail of this data.

My research design also requires me to link deforestation to rivers. Any sediments rivers would follow the flow of water through the terrain. Starting from drainage trenches, the flow of water accumulates until the stream can be considered a river. Surrounding any river there is therefore a watershed area which feeds into it. To model this, I split rivers into segments of 1km and then use the methodology of @bartos_2020 with the digital elevation model provided by @europeanspaceagencyCopernicus2022 to calculate exact drainage polygons around the river segments. This approach requires extensive data pre-processing and matching algorithms which I further elaborate on in @sec-appendix-hydro. These polygons are then used to extract deforestation as discussed above. In @#fig-drainage-extraction I present an example of the detailed results of this process.

@@

The Brazilian National Water Agency (ANA) collects standardized information on water quality in the @agencianacionaldeaguasanaPortalQualidadeAguas2024. I download all available historical data from the HidroWeb system of the National Water Resources Information System (SNIRH) and aggregate them by year, resulting in an unbalanced panel of 4,522 stations in 45,094 data points. In my analysis I focus on the variables of pH, turbidity, biochemical oxygen demand (BOD5.20), dissolved oxygen, total residue, total nitrogen, and nitrate concentration. Not all variables are measured at all times, please refer to @tbl-summary-sensor-data for summary statistics. Maps of the spatial distribution of all stations across time intervals are provided in @fig-stations. Before the year 2000 only few stations mostly situated in the urban South-West report measuruments. Therefore I choose to drop earlier observations and focus on the later years with more complete coverage.

@@

To account for potential confounding factors, several control variables are included in the analysis. These controls are GDP per capita, average schooling grades, a vaccination index, primary care coverage, doctors per capita, urbanization and the availability of sewage systems. Data for GDP per capita are sourced from @ibgeProdutoInternoBruto1996. Average schooling grades in form of the IDEB index are sourced @institutonacionaldeestudosepesquisaseducacionaisinepIndiceDesenvolvimentoEducacao2005. The vaccination index, representing the coverage of essential immunizations, is derived from health department records @ministeriodasaudemsSistemaImunizacoes1994.^[I focus on the number of administered vaccinations against yellow fever, hepatitis A and B, polio, and Haemophilus influenzae type B. In the analysis I use a rolling 5-year average of annual vaccination rates of newborns.] I obtain the coverage with primary care and the number of doctors per capita from @ministeriodasaudemsCoberturaAtencaoBasica2007. Finally, data on urbanization and sewage availability are obtained from municipal health and infrastructure reports @ministeriododesenvolvimentoregionalmdrSistemaNacionalInformacoes1995. Data on migration and race are obtained from @ibgeCensoDemografico2000, the waves of 2000 and 2010. To be specific, I utilize question V0417 (*NASCEU NESTE MUNICÍPIO?*, V0618 in 2010) to identify migrants and item V0408 (V0606 in 2010) to identify race. Census sample weights are applied during aggregation by municipality.

@@

If deforestation was to expose populations to health outcomes along rivers, one should be able to estimate an impact on sensor-measured river pollution. Evidence from environmental science shows that soil erosion increases turbidity by introducing sediments into water bodies, reducing light penetration and affecting aquatic life [@u.s.epaTurbidityParameterFactsheet2021]. Eroded soils carry organic matter and nutrients, raising biochemical oxygen demand levels and potentially depleting dissolved oxygen, crucial for aquatic organisms [@u.s.epaDissolvedOxygenParameter2021]. The increased runoff also contributes to higher total residues, nitrogen, and nitrate concentrations, leading to harmful algal blooms [@u.s.epaNutrientsParameterFactsheet2021]. Altered pH levels can further disrupt the ecological balance and water quality [@u.s.epaPHParameterFactsheet2021].

In @tbl-reg-deforestation-sensors I present regression results for the corellation between aggregate upstream deforestation and pollution.^[See @tbl-reg-deforestation-sensors-sd for standardized coefficients.] I find that deforestation of ten percent of upstream area is associated with a increase in the water's pH level by 0.12 or 0.16 standard deviations, an increase in total residue by 11.55 or 0.4 standard deviations and a decrease of biochemical oxygen demand by 1.86 or 0.3 standard deviations c.p.. These coefficients are statistically significant at the 5%, 10% and 5% level respectively. The increases in total residue and pH match expectations while the decrease in biochemical oxygen demand is surprising. Coefficients for turbidity and dissolved oxygen are negative but statistically insignificant. Nitrate and total nitrogen readings are on average insignificantly increased.

@@

To further shed a light on the mechanism of impact behind this correlation I now turn to the effect of the different replacing land cover types. For any deforestation event, I identify what forests were replaced by and group this into four categories: Pasture, agriculture, urban environments, and mining. In @tbl-reg-deforestation-sensors-replacement I present OLS results of this analysis, with otherwise equivalent model parameters to the above. I find that deforestation for pasture is associated with increased turbidity and BOD5.20 measurements. Observed pH, dissolved oxygen, total residue, total nitrogen and nitrate levels are decreased but the coefficients are not statistically significant. In contrast to that, increased agricultural area is correlated with decreased turbidity and otherwise similar insignificant coefficients. When urban areas replace forests, I observe an on average reduced level of BOD5.20. Coefficients for pH, total residue, nitrogen and nitrates are positive but not statistically different from zero. In cases where mining displaces forests I find strong increases in turbidity and BOD5.20. Readings of dissolved oxygen are on average lower. Estimates for total residue, nitrogen and nitrates are positive but high variation does not allow for confidence in the latter estimates.

@@

In @tbl-reg-deforestation-sensors-replacement-levels I present results when regressing on levels of land cover instead of deforestation events. This model is more intuitive to interpret as I am pivoting away from only looking at year-to-year changes in land cover and sensor measurements to also considering the *stock* of land cover present upstream of a sensor. However, this approach is likely to suffer from spurious variation. The decreasing trend in cumulative forest cover (and the increasing trends in other land cover categories) might coincide with trends in unobservables, yielding biased coefficients. Overall, this analysis yields similar results. In contrast to my main findings, coefficients for pH on pasture and agriculture are positive. High agriculture levels are associated with high instead of lower levels of turbidity. The negative coefficients for the relationship between pasture, agriculture, and dissolved oxygen as well as total residue are now statistically significant. Nitrate coefficients for pasture, agriculture and urban land cover are now significant. Finally, the relationships between mining and turbidity, BOD5.20, and dissolved oxygen are not statistically significant any more. 
Finally, in @tbl-reg-deforestation-sensors-distance I observe how the effect of deforestation differs with the distance to the deforestation. To this end I regress the pollution indicators on deforestation and its interaction with the distance bin it is measured upstream from the station. All effects with the exception of turbidity decrease in distance although point estimates for the interaction term are significant only for dissolved oxygen and total residue.

In conclusion, I find several theoretical results validated. Deforestation is associated with altered pH and total residue in the water. Unexpectedly, biochemical oxygen demand is on average lower when there is upstream deforestation. It turns out that mining is most strongly correlated with sensor readings. I find increased turbidity, BOD5.20 and decreased dissolved oxygen in water downstream of mines. Surprinsingly, I do not find increased nitrogen and nitrate readings when there is increased upstream farming activity. However, turbidity and BOD5.20 levels are increased when pasture areas are created. Many other idiosyncratic factors influence the discussed relationship. While the fixed effects approach alleviates some concern, I want to stress that I do not claim any of these results to be causal relationships.

## Health Outcomes {#sec-results-health}

@@

Before I turn to the results of the analysis, I present first stage results. In @tbl-reg-deter-first-stage I regress upstream DETER cloud cover on upstream deforestation for all municipalities. I present three different specifications, each consecutively adding control variables. Since data on doctors and primary care coverage is only available from 2010 onwards, the last model is estimated on a restricted sample. I find that in all specifications cloud coverage is negatively correlated with forest cover. A 10 percentage point increase in cloud coverage is associated with an increase in annual upstream deforestation of between .03 and .05 percentage points or about 0.1 standard deviations. All coefficients are statistically significant at all conventional levels.
